---
title: "microcourse"
output: html_document
date: "2024-05-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
^Hey, what's all this about? Well echo = TRUE means if you knit this document (make it ready for sharing), all the code AND the results will be included in the document.

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

These are called **chunk options**. There so *so* many. Here are some of the most relevant: 

Chunk option    | Description
----------------|--------------
include = FALSE | Prevents code and results from appearing in the finished file.
echo = FALSE    | Prevents code but not results from appearing in the finished file.
message = FALSE | Prevents messages that are generated by code from appearing in the finished file.
warning = FALSE | Prevents warnings that are generated by code from appearing in the finished.
fig.cap = "..." | Adds a caption to any graphical results.


When you put the chunk option code in the initial r setup chunk, it applies the option globally throughout this R Markdown file. ~~You can also just delete it, which is what most people do~~ 

The {r setup} part just names the chunk. So {r cars} for example names the chunk 'cars'. This helps you organize your code chunk by chunk ~~but a lot of people don't use it~~

```{r cars}
summary(cars)
```

Otherwise, welcome to R Markdown! For more details on using R Markdown see <http://rmarkdown.rstudio.com>. Markdown provides simple formatting syntax to easily create coding files that can be made into HTML, PDF, or Word docs. If you click **Knit** up above, a finished file in your selected file format will be created. 


# BasicBasics






